A Shell Model for Real-Time Simulation of Intra-ocular Implant Deployment

### 2009-11-18 ISBMS09/paper.tex
Final version
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@69 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Although various shapes of intra-ocular lens implants exist they all share common properties: they are three-dimensional structures, usually made of acrylic
%\footnote{acrylic is the common name for polymethylmethacrylate} 
material, and are much thinner in one direction than in the two other ones. This last characteristic is what classifies these lenses as thin objects, i.e. three-dimensional objects that can be abstracted as two-dimensional structures. Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thorough representations. Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects. Most of them fall into one of the following two categories: plate theory or shell theory. Those theories have been a subject of interest in the mechanical community for decades. The difference between these two kinds of structures is very well explained by Liu \textit{et al.}~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations. 
%For instance, if we consider the horizontal board of a bookshelf, that board can be approximated as a plate structure and the transversal loads are the weight of the books. A typical deformation of the board is illustrated in figure~\ref{fig-boards}(a). Conversely, a shell structure can carry loads in all directions, and therefore undergoes bending, twisting and in-plane deformation (see Figure~\ref{fig-boards}(b)). 
Considering the folded shape of the implant within the injection device (see Figure \ref{fig-surgery}), a thin plate model would not correctly capture the deformation and therefore a shell formulation was retained to model the intra-ocular lens behaviour.

%\begin{figure}[ht]
%\begin{center}
%\includegraphics[width=4.5cm]{images/intraocular_implant_1}
%\hspace{1.5cm}
%\includegraphics[width=4.5cm]{images/intraocular_implant_2}
%\caption{Different types of lens implants. All share similar properties: they are three-dimensional structures, usually made of acrylic material, and are much thinner in one direction than in the two other ones.}
%\label{fig-implants}
%\end{center}
%\end{figure}

%\begin{figure}[ht]
%\begin{center}
%\includegraphics[width=5.5cm]{images/board_bending}
%\hfill
%\includegraphics[width=5.2cm]{images/board_twist}
%\caption{Illustration of a key difference between different models of thin objects. While thin plate theory allows to describe bending (left), it cannot represent more complex deformations such as twist (right) which is captured by shell theory.}
%\label{fig-boards}
%\end{center}
%\end{figure}

@@

In this paper we consider the problem of lens implant folding, unfolding and deployment within a very constrained space. An important difference with previous works is that the ratio stiffness / mass of artificial lenses is very high compared to many material (such as fabric), resulting in added difficulties to obtain a numerically stable simulation. The closest work to what we propose is the one of Thomaszewski \emph{et al.} \cite{Thomaszewski06} who derive their formulation from thin shell analysis in a co-rotational framework. However, the authors use very soft and deformable material (Young's modulus $E = 5000\,Pa$ and Poisson's ratio $\nu = 0.25$ while the stiffness of intra-ocular implants are three orders of magnitude higher).

@@

\begin{figure}
\centering
\includegraphics[height=3cm]{images/triangle_membrane}
\hspace{1cm}
\includegraphics[height=3.25cm]{images/triangle_bending}
\caption {A triangular shell element can be defined as a combination of a triangular in-plane membrane element (left) and a triangular thin plate in bending (right). The different degrees of freedom $v$ and $u$ of both models are illustrated above.}
\label{fig-triangle}
\end{figure}

@@

where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour. In the simple case of Hooke's law we have:

@@

We can calculate the strains from the flat-plate theory using $  e_{xx} = -z \frac{\partial^2u_z}{\partial x^2} $, $ e_{yy} = -z \frac{\partial^2u_z}{\partial y^2} $, and $ e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y} $.  
Hence, using the above equations and equation (\ref{eq-deflection}), we have
\begin{equation}
\begin{bmatrix}
e_{xx} \\
e_{yy} \\
e_{xy}
\end{bmatrix}
= 
-z
\begin{bmatrix}
0 & 0 & 0 & 2 & 0 & 0 & 6x & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 2 & 0 & 2x & 6y \\
0 & 0 & 0 & 0 & 2 & 0 & 0 & 4y & 0 \\
\end{bmatrix}
\textbf{c}
\label{eq-strains}
\end{equation} 
or symbolically $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ stands for the $3\times 9$ matrix in equation \ref{eq-strains}, including the pre-multiplying constant $-z$. Noting from equation (\ref{eq-U}) that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
\begin{equation}
\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu}
\end{equation} 
where $\textbf{b} = \textbf{DC}^{-1}$. 
Knowing that the stiffness matrix $\textbf{K}_e$ for an element is obtained from
\begin{equation}
\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV
\end{equation} 
where $\boldsymbol\chi$ is the material matrix, the substitution of $\textbf{b}$ into this expression yields
\begin{equation}
\textbf{K}_e = (\textbf{C}^{-1})^T \int_v \textbf{D}^{T} \boldsymbol\chi \textbf{D} dV \textbf{C}^{-1}
\end{equation} 
The integration is carried out numerically using Gauss points located at the middle of each edge of the triangle.

@@

\begin{table}[ht]
	\centering
	\begin{tabular}{p{8.5cm}|c|c|}
	\cline{2-3}
	\multirow{5}{*}{\includegraphics[height=2.75cm]{images/clamped_square}} & $q$ & $z_{max}$ \tabularnewline
	\cline{2-3}
	& $\,1\,$ & $\, 0.1218 \,$ \tabularnewline
	& $\,2\,$ & $\, 0.2475 \,$ \tabularnewline		
	& $\,3\,$ & $\, 0.3747 \,$ \tabularnewline	
	& $\,4\,$ & $\, 0.5050 \,$ \tabularnewline		
	& $\,5\,$ & $\, 0.6374 \,$ \tabularnewline
	\cline{2-3}
	\end{tabular}
	\caption{Comparison of our shell model with theoretical results on the bending of a square plate. An error of less than $1\,\%$ was found between our simulation and the theoretical results.}
	\label{tab-results}
\end{table}

@@

Cataract surgery consists in three main steps: capsulorhexis, phacoemulsification, and implantation of an intra-ocular lens. Capsulorhexis is the technique used to remove a part of the anterior lens capsule. Phacoemulsification consists in using a surgical device which tip vibrates at an ultrasonic frequency to emulsify the natural lens material and then aspirate the  fragments of the cortical material. After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about 2 mm) using a foldable intra-ocular lens (see figure \ref{fig-surgery}). The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification.

@@

To simulate the insertion and deployment of an intra-ocular lens, we first created a triangulation of the lens surface. Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (figure \ref{fig-implant}). The lens mesh contains $743$ triangles and $473$ nodes. Models of the injection device and the entire eye anatomy were also created. Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented into the table \ref{tab-parameters}.

@@

To obtain the folded geometry of the lens we simulated the folding process by first folding the hooks (also called \emph{haptics}) onto the implant body. The body was then bent while keeping the hooks inside to obtain the shape described in figure \ref{fig-implant}. The folded implant was then placed into the injection device. The simulation of the injection consists in pushing the intra-ocular implant within the injection device into the lens capsule. During these stages of the simulation, complex contacts occur and consist of self collisions of the lens as well as collisions between the lens and the injector and later with the capsule. To solve the contacts we use the contact warping method proposed Saupin \emph{et al.} \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models.

\begin{figure}[!h]
\centering
\includegraphics[height=3.5cm]{images/IOL}
\hfill
\includegraphics[height=3.5cm]{images/mesh_implant}
\hfill
\includegraphics[height=2.5cm]{images/implant_haptics_in}
\hfill
\includegraphics[height=3.0cm]{images/implant_folded}
\caption [Lens implant and its mesh] {An actual intra-ocular implant and the triangular mesh used in our simulations. Notice the higher density of elements in areas where large deformations will take place.}
\label{fig-implant}
\end{figure}

Results of our simulation are illustrated in figure \ref{fig-simu-results}. We can notice the progressive deployment of the implant when it exits the injector.  The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule. Due to the high stiffness and low mass of the lens, a direct solver was used at each time step rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).

\begin{figure}[!h]
\centering
\includegraphics[height=3.5cm]{images/simu_injection_step_1}
\hfill
\includegraphics[height=3.5cm]{images/simu_injection_step_2}
\hfill
\includegraphics[height=3.5cm]{images/simu_injection_step_3}
\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule.}
\label{fig-simu-results}
\end{figure}

@@

We proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach. This simple shell element can efficiently handle both membrane, bending and twist forces. The validity of our approach has been demonstrated though comparison with theoretical results. We then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation. These preliminary results are very encouraging, and show the potential of such models. Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels.

\section*{Acknowledgments}
We are very grateful to J\'er\'emy Ringard for his invaluable help in improving the eye anatomical model, to J\'er\'emie Dequidt and Damien Marchal for understanding Blender, to Fr\'ederick Roy for his help with SOFA, and Nadia Boubchir for her persistence in motivating us to obtain these results in time...

>>>>

Although various shapes of intra-ocular lens implants exist they all share common properties: they are three-dimensional structures, usually made of acrylic
%\footnote{acrylic is the common name for polymethylmethacrylate} 
material, and are much thinner in one direction ($0.13$\,mm) than in the two other ones ($6$ and $13$\,mm). This last characteristic is what classifies these lenses as thin objects, i.e. three-dimensional objects that can be abstracted as two-dimensional structures. Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thorough representations. Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects. Most of them fall into one of the following two categories: plate theory or shell theory. Those theories have been a subject of interest in the mechanical community for decades. The difference between these two kinds of structures is very well explained by Liu \textit{et al.}~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations. 
For instance, if we consider the horizontal board of a bookshelf, that board can be approximated as a plate structure and the transversal loads are the weight of the books. A typical deformation of the board is illustrated in figure~\ref{fig-boards} left. Conversely, a shell structure can carry loads in all directions, and therefore can undergo bending, twisting and in-plane deformation (see Figure~\ref{fig-boards} right). 
Considering the folded shape of the implant within the injection device (see Figure \ref{fig-surgery}), a thin plate model would not correctly capture the deformation and therefore a shell formulation was retained to model the intra-ocular lens behaviour.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=6cm]{images/board_bending}
\hfill
\includegraphics[width=6cm]{images/board_twist}
\caption{Illustration of a key difference between various models of thin objects. While thin plate theory allows to describe bending (left), it cannot represent more complex deformations such as twist (right) which is captured by shell theory.}
\label{fig-boards}
\end{center}
\end{figure}

@@

In this paper we consider the problem of lens implant folding, unfolding and deployment within a very constrained space. An important difference with previous works is that the ratio stiffness / mass of artificial lenses is very high compared to many material (such as fabric), resulting in added difficulties to obtain a numerically stable simulation. Indeed, if the mass density of cotton and acrylic is about the same, the stiffness of intra-ocular implants is three orders of magnitude higher. The closest work to what we propose is the one of Thomaszewski \emph{et al.} \cite{Thomaszewski06} who derive their formulation from thin shell analysis in a co-rotational framework. However, the authors use very soft and deformable material (Young's modulus $E = 5000\,Pa$ and Poisson's ratio $\nu = 0.25$).

@@

\begin{figure}
\centering
\includegraphics[height=3cm]{images/triangle_membrane}
\includegraphics[height=3.5cm]{images/triangle_bending}
\caption {A triangular shell element can be defined as a combination of a triangular in-plane membrane element (left) and a triangular thin plate in bending (right). The different degrees of freedom $v$ and $u$ of both models are illustrated above.}
\label{fig-triangle}
\end{figure}

@@

where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour. The implant is very stiff and we therefore assume that the local deformations remain limited during the deployment and a linear constitutive law is sufficient. Thus in the simple case of Hooke's law we have:

@@

We can calculate the strains from the flat-plate theory using:
\begin{equation}
e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
\end{equation} 
\begin{equation}
e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
\end{equation} 
\begin{equation}
e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
\end{equation}

Hence, using the above equations and equation (\ref{eq-deflection}), we have
\begin{equation}
\begin{bmatrix}
e_{xx} \\
e_{yy} \\
e_{xy}
\end{bmatrix}
= 
-z
\begin{bmatrix}
0 & 0 & 0 & 2 & 0 & 0 & 6x & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 2 & 0 & 2x & 6y \\
0 & 0 & 0 & 0 & 2 & 0 & 0 & 4y & 0 \\
\end{bmatrix}
\textbf{c}
\label{eq-strains}
\end{equation} 
or symbolically $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ stands for the $3\times 9$ matrix in equation \ref{eq-strains}, including the pre-multiplying constant $-z$. Noting from equation (\ref{eq-U}) that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
\begin{equation}
\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu}
\end{equation} 
where $\textbf{b} = \textbf{DC}^{-1}$. 
Knowing that the stiffness matrix $\textbf{K}_e$ for an element is obtained from
\begin{equation}
\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV
\end{equation} 
where $\boldsymbol\chi$ is the material matrix, the substitution of $\textbf{b}$ into this expression yields
\begin{equation}
\textbf{K}_e = (\textbf{C}^{-1})^T \int_v \textbf{D}^{T} \boldsymbol\chi \textbf{D} dV \textbf{C}^{-1}
\end{equation} 
The integration is carried out numerically using Gauss points located at the middle of each edge of the triangle.

@@

\begin{table}[ht]
	\centering
	\begin{tabular}{p{9cm}|c|c|}
	\cline{2-3}
	\multirow{5}{*}{\includegraphics[height=3.5cm]{images/clamped_square}} & $q$ & $z_{max}$ \tabularnewline
	\cline{2-3}
	& $\,1\,$ & $\, 0.1218 \,$ \tabularnewline
	& $\,2\,$ & $\, 0.2475 \,$ \tabularnewline		
	& $\,3\,$ & $\, 0.3747 \,$ \tabularnewline	
	& $\,4\,$ & $\, 0.5050 \,$ \tabularnewline		
	& $\,5\,$ & $\, 0.6374 \,$ \tabularnewline
	\cline{2-3}
	\end{tabular}
	\vspace{1cm}
	\caption{Comparison of our shell model with theoretical results on the bending of a square plate. An error of less than $1\,\%$ was found between our simulation and theoretical results.}
	\label{tab-results}
\end{table}

@@

Cataract surgery consists in three main steps: capsulorhexis, phacoemulsification, and implantation of an intra-ocular lens. Prior to starting the surgery, a viscoelastic fluid is introduced into the capsule to facilitate capsulorhexis creation and provide protection during phacoemulsification. This fluid remains in the capsule for the duration of the surgery, including the injection of the implant. Capsulorhexis is the technique used to remove a part of the anterior lens capsule. Phacoemulsification consists in using a surgical device which tip vibrates at an ultrasonic frequency to emulsify the natural lens material and then aspirate the  fragments of the cortical material. After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about 2 mm) using a foldable intra-ocular lens (see figure \ref{fig-surgery}). The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification. In some cases the implant can be flipped or  the hooks (also called \emph{haptics}) can break. Therefore the simulation of the insertion and deployment of the implant is crucial for achieving a successful surgery.

@@

To simulate the insertion and deployment of an intra-ocular lens, we first created a triangulation of the lens surface. Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (figure \ref{fig-mesh}). This was done by noting the constraints applied by the surgeon to the haptics while inserting the implant within the injection device. During this stage, the haptics are folded onto the implant body, leading to high stresses at the junctions. The lens mesh contains $743$ triangles and $473$ nodes. Models of the injection device and the entire eye anatomy were also created. Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented in table \ref{tab-parameters}.

@@

The first difficulty is to obtain the folded geometry of the lens within the injection device. This step is not important for the training process and does not need to be interactive. Indeed the surgeon does not always have to prepare the implant as some injection devices are readily available with a folded implant already in place. We simulate the folding process by first folding the haptics onto the implant body. The body was then bent while keeping the haptics inside to obtain the shape described in figure \ref{fig-implantFolding}. The whole process was carried out by applying the necessary forces and boundary conditions on the body and haptics of the implant. The folded implant was then placed into the injection device. The simulation of the injection consists in pushing the intra-ocular implant within the injection device into the lens capsule. During these stages of the simulation, complex contacts occur and consist of self collisions of the lens as well as collisions between the lens and the injector and later with the capsule. To solve the contacts we use the contact warping method proposed Saupin \emph{et al.} \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models. As adhesions between the haptics and the body is often observed in surgery, friction is also taken into account in the contact response process.

\begin{figure}[!h]
\centering
\includegraphics[height=5cm]{images/IOL}
\includegraphics[height=5cm]{images/mesh_implant}
\caption [Lens implant and its mesh] {An actual intra-ocular implant and the triangular mesh used in our simulations. Notice the higher density of elements in areas where large deformations will take place.}
\label{fig-mesh}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=4cm]{images/implant_folding1}
\includegraphics[width=3.5cm]{images/implant_folding2} \\
\includegraphics[width=3cm]{images/implant_folding3}
\includegraphics[width=5.5cm]{images/implant_folding4}
\caption [Folding of intra-ocular implant] {Top: intermediary steps of the intra-ocular implant folding. Bottom: fully folded implant ready to be placed into the injection device.}
\label{fig-implantFolding}
\end{figure}

Results of our simulation are illustrated in figure \ref{fig-simu-results}. We can notice the progressive deployment of the implant when it exits the injector.  The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule. Due to the high stiffness and low mass of the lens, a direct sparse solver was used at each time step (dt $= 0.01$\,s) rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).

\begin{figure}[!h]
\centering
\includegraphics[width=4cm]{images/surgery1}
\hfill
\includegraphics[width=4cm]{images/surgery2}
\hfill
\includegraphics[width=4cm]{images/surgery3} \\
\vspace{0.1cm}
\includegraphics[width=4cm]{images/simu1}
\hfill
\includegraphics[width=4cm]{images/simu2}
\hfill
\includegraphics[width=4cm]{images/simu3}
\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule. Top: images from a real cataract surgery, courtesy of Dr. Tarek Youssef, ophtalmologist in Canada. Below: our simulation of the implant's deployment.}
\label{fig-simu-results}
\end{figure}

@@

We proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach. This simple shell element can efficiently handle both membrane, bending and twist forces. The validity of our approach has been demonstrated though comparison with theoretical results. It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation. These preliminary results are very encouraging and show the potential of such models. Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels. Regarding improvements of our current model and simulation, we plan to account for the interaction with the viscoelastic fluid (currently approximated by a high damping). Computation times also need to be improved through the use of more optimised meshes and by porting our parallelisable algorithm to Graphics Processing Units.

\section*{Acknowledgments}
We are very grateful to J\'er\'emy Ringard for his invaluable help in improving the eye anatomical model, to J\'er\'emie Dequidt and Damien Marchal for understanding Blender, to Fr\'ederick Roy for his help with SOFA, and Nadia Boubchir for her persistence in motivating us to obtain these results in time.

\bigskip

We would also like to thank INRIA and the SOFA project, as well as CSIRO Preventative Health National Research Flagship, ICTC, The Australian e-Health Research Centre for their fundings.
